(Re-)bridging the anatomy ontologies with SSSOM

### 2024-06-16 paper.tex
Rename the paper source file.
This is to prepare the merge with the presentation repository.

While simple, the cross-reference method has several limitations. (i) It
does not allow to record any metadata about the mapping, such as: who
asserted that the two terms should be mapped? On what basis? When was
the mapping reviewed?  (ii) The method does not allow for any nuance:
either two terms are mapped or they are not, without room to express
more subtle relations. (iii) Cross-references have no clear and
universally agreed upon meaning. When two terms are mapped to each
other, the meaning of the mapping is left unspecified, and typically has
to be inferred from the ontologies that are being mapped (for example, a
mapping between the species-neutral Uberon and the
\species{Drosophila}-specific FBbt can be inferred to be a cross-species
mapping). This is made worse by the fact that cross-references in OBO
ontologies are used for many different things beyond just mappings; for
example, when used to annotate a term definition, cross-references
typically attribute a source for the definition, akin to a citation in
an academic paper.

@@

To construct a consistent and maintainable cross-species anatomy
ontology, we have updated the integration mechanism between Uberon, CL,
and the species-centric ontologies by using SSSOM to manage the mappings
betweeen the ontologies. While full transition to SSSOM is not yet
complete, we now have the needed infrastructure in place to complete
this migration, and we have a more robust and more flexible pipeline for
creating the bridges needed for \emph{composite-metazoan}. The logic for
deriving the bridges is expressed using a simple DSL (SSSOM/T-OWL) that
most ontology editors should be able to master quickly, should the need
arise to modify said logic (for example to add a new bridge). While the
precise SSSOM/T-OWL rules we use are obviously specific to the needs of
Uberon, we believe the general approach of maintaining mappings as SSSOM
sets and transforming them into OWL bridges using a simple DSL could be
used by any project that requires integrating several ontologies
together.

>>>>

While simple, the cross-reference method has several limitations. (i) It
does not allow to record any metadata about the mapping, such as: who
asserted that the two terms should be mapped? On what basis? When was
the mapping reviewed?  (ii) The method does not allow for any nuance:
either two terms are mapped or they are not, without room to express
more subtle relations. (iii) Cross-references are embedded within the
ontology itself, and are therefore not easily reusable by third parties.
(iv) Cross-references have no clear and universally agreed upon meaning.
When two terms are mapped to each other, the meaning of the mapping is
left unspecified, and typically has to be inferred from the ontologies
that are being mapped (for example, a mapping between the
species-neutral Uberon and the \species{Drosophila}-specific FBbt can be
inferred to be a cross-species mapping). This is made worse by the fact
that cross-references in OBO ontologies are used for many different
things beyond just mappings; for example, when used to annotate a term
definition, cross-references typically attribute a source for the
definition, akin to a citation in an academic paper.

@@

In this project, we have (i) expanded the SEMAPV vocabulary so that
cross-species mappings can be specifically represented in SSSOM; (ii)
introduced a domain-specific language (SSSOM/T-OWL) to allow converting
SSSOM mappings into arbitrary OWL axioms; (iii) implemented said
language in a new ROBOT plugin to inject SSSOM-derived axioms into a OWL
ontology; (iv) used those new tools to update the integration mechanism
between Uberon, CL, and the species-centric ontologies, with an approach
centered on the use of SSSOM, rather than cross-references, to manage
the mappings between the ontologies.

As a result, cross-species mappings across anatomy ontologies are now
available as a \emph{bona fide} release artefact of Uberon, provided as
a distinct file in a known location in a standard format, thereby
greatly facilitating the reuse of said mappings by any interested third
party. The logic for deriving the bridge files (needed to correctly
merge Uberon and the species-specific ontologies), now expressed in the
SSSOM/T-OWL language, is itself consequently more easily maintainable
and reusable.

While several aspects of this project (most notably the precise
SSSOM/T-OWL rules we use) are specific to the needs of Uberon, we
believe the general approach of maintaining mappings as SSSOM sets and
transforming them into OWL bridges using a simple DSL could be
generalised to any project that requires integrating several ontologies
together.
